---
title: "Setting Up a SWAT+ Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting Up a SWAT+ Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette walks through every step involved in setting up and running a
SWAT+ simulation using the **swatplusEditoR** package.
The workflow assumes that watershed delineation and HRU definition have already
been completed—for example using the [rQSWATPlus](https://github.com/celray/rQSWATPlus)
package—and that a project database (`.sqlite` or `.db`) already exists.

The typical simulation setup follows these steps:

1. **Initialise the project** – create a project database from delineated
   watershed data using rQSWATPlus (or load an existing database)
2. **Load and inspect the project** – validate the database and review GIS
   tables
3. **Add weather data** – supply weather station observations and/or weather
   generator parameters
4. **Configure simulation parameters** – set the time period, output options,
   and adjust model parameters
5. **Write SWAT+ input files** – generate the complete TxtInOut file set
   entirely within R
6. **Run the model** – execute the bundled SWAT+ executable

---

## Step 1: Initialise the Project with rQSWATPlus

The first step is to delineate the watershed and define HRUs.
The **rQSWATPlus** package automates this process from DEM, land-use, and soil
rasters.  The example below uses the Ravn catchment data that ships with
rQSWATPlus:
  
```{r rqswatplus-init}
# ── Locate example datasets bundled with rQSWATPlus ──────────────────────────
dem         <- system.file("extdata", "ravn_dem.tif",      package = "rQSWATPlus")
landuse     <- system.file("extdata", "ravn_landuse.tif",  package = "rQSWATPlus")
soil        <- system.file("extdata", "ravn_soil.tif",     package = "rQSWATPlus")
lu_lookup   <- system.file("extdata", "ravn_landuse.csv",  package = "rQSWATPlus")
soil_lookup <- system.file("extdata", "ravn_soil.csv",     package = "rQSWATPlus")
outlet      <- system.file("extdata", "ravn_outlet.shp",   package = "rQSWATPlus")

# ── Run the complete delineation + HRU workflow ──────────────────────────────
project <- rQSWATPlus::qswat_run(
  project_dir     = file.path(tempdir(), "ravn_quick"),
  dem_file        = dem,
  landuse_file    = landuse,
  soil_file       = soil,
  landuse_lookup  = lu_lookup,
  soil_lookup     = soil_lookup,
  outlet_file     = outlet,
  threshold       = 500,
  slope_breaks    = c(0, 5, 15, 9999),
  landuse_threshold = 5,
  soil_threshold    = 5,
  db_file         = "swat.db",
  quiet           = TRUE
)
```

After `qswat_run()` completes, the `project` object contains the paths to all
generated files and, crucially, the `db_file` field pointing to the SQLite
project database.

> **Tip:** If you already have an existing SWAT+ Editor project database you
> can skip this step and simply point `project$db_file` at that file.

---

## Step 2: Load and Inspect the Project

Once the database exists, use **swatplusEditoR** to load and validate it:

```{r load-project}
library(swatplusEditoR)

# Load the project – validates the database and lists available GIS tables
project <- load_project(project)
```

### Reviewing Project Information

```{r project-info}
# Summary of project status
info <- get_project_info(project)
info
```

`get_project_info()` returns a list with counts for subbasins, channels, HRUs,
weather stations, and timestamps showing when files were last written or the
model was last run.

### Reading GIS Data

You can inspect any of the GIS tables directly:

```{r gis-data}
# Read all GIS tables at once
gis <- read_gis_data(project)
names(gis)

# Or read individual tables
subbasins <- read_gis_subbasins(project)
head(subbasins)

channels <- read_gis_channels(project)
head(channels)

hrus <- read_gis_hrus(project)
head(hrus)
```

Other GIS readers include `read_gis_lsus()`, `read_gis_aquifers()`,
`read_gis_routing()`, `read_gis_water()`, and `read_gis_points()`.

---

## Step 3: Add Weather Data

SWAT+ requires weather forcing data.  You can add **weather stations** (which
reference `.cli` files on disk) and **weather generators** (monthly statistics
used by the built-in weather generator).

### Adding Weather Stations

Create a data frame where each row represents a station.  The `pcp`, `tmp`,
`slr`, `hmd`, `wnd`, and `pet` columns contain the file names of the
corresponding climate input files:

```{r add-stations}
stations <- data.frame(
  name     = c("station1", "station2"),
  lat      = c(55.8, 55.9),
  lon      = c(9.6, 9.7),
  pcp      = c("pcp1.cli", "pcp2.cli"),
  tmp      = c("tmp1.cli", "tmp2.cli"),
  slr      = c("slr1.cli", "slr2.cli"),
  hmd      = c("hmd1.cli", "hmd2.cli"),
  wnd      = c("wnd1.cli", "wnd2.cli"),
  pet      = c("pet1.cli", "pet2.cli"),
  atmo_dep = c("atmo1.cli", "atmo2.cli"),
  stringsAsFactors = FALSE
)

add_weather_stations(project, stations)
```

### Setting the Weather Data Directory

Tell SWAT+ where to find the climate files on disk:

```{r set-weather-dir}
set_weather_dir(project, "/path/to/weather/data")
```

### Listing and Updating Stations

```{r list-stations}
# View all registered stations
list_weather_stations(project)

# Update a single station field (e.g. change the PCP file)
update_weather_station(project, station_id = 1, pcp = "pcp1_v2.cli")
```

### Adding Weather Generators

Weather generator (WGN) parameters provide monthly climate statistics that
SWAT+ uses when observed data are missing:

```{r add-wgn}
wgn <- data.frame(
  name     = "wgn_ravn",
  lat      = 55.85,
  lon      = 9.65,
  elev     = 50,
  rain_yrs = 30,
  stringsAsFactors = FALSE
)

add_weather_generators(project, wgn)
```

You can optionally supply a `monthly_data` data frame with 12 rows of monthly
statistics (temperature means/std dev, precipitation, solar radiation, etc.).

---

## Step 4: Configure the Simulation

### Setting the Simulation Period

Define the start and end dates for the model run.  Julian days are used
(1 = January 1, 365 = December 31):
  
```{r sim-time}
set_simulation_time(
  project,
  day_start = 1,
  yrc_start = 2000,
  day_end   = 365,
  yrc_end   = 2010,
  step      = 0           # 0 = daily time step
)
```

### Configuring Output Options

Control what SWAT+ writes during and after the simulation:

```{r print-options}
set_print_options(
  project,
  nyskip  = 2,           # skip 2 years for warm-up
  csvout  = TRUE,        # write CSV output files
  dbout   = FALSE,       # do not write SQLite output
  cdfout  = FALSE        # do not write NetCDF output
)
```

### Updating Model Parameters

`update_parameters()` is a general-purpose function that can modify any
database table.  Common tables include:

| Table | Contents |
|-------|----------|
| `hydrology_hyd` | Curve number, percolation, lateral flow |
| `topography_hyd` | Slope, slope length |
| `aquifer_aqu` | Groundwater parameters |
| `channel_cha` | Channel geometry and erodibility |
| `codes_bsn` | Basin-wide model codes |
| `parameters_bsn` | Basin-wide calibration parameters |

```{r update-params}
# Update curve number for ALL HRUs
update_parameters(project, "hydrology_hyd", list(cn2 = 65))

# Update only specific HRUs (by row ID)
update_parameters(project, "hydrology_hyd", list(cn2 = 70), ids = c(1, 2, 3))

# Adjust percolation coefficient
update_parameters(project, "hydrology_hyd", list(perco = 0.5))

# Modify aquifer parameters
update_parameters(project, "aquifer_aqu", list(gw_dp = 30, bf_max = 1.5))
```

### Configuring GWFLOW (Optional)

If your model requires a detailed groundwater module:

```{r gwflow}
# Check current GWFLOW status
get_gwflow_status(project)

# Initialise GWFLOW grid
init_gwflow(project, cell_size = 200, row_count = 100, col_count = 150)

# Get/update base configuration
base <- get_gwflow_base(project)
update_gwflow_base(project, boundary_conditions = "constant_head")

# Get/update zones
zones <- get_gwflow_zones(project)
update_gwflow_zones(project, zone_id = 1, aquifer_k = 15.0)
```

---

## Step 5: Write SWAT+ Input Files

With the database fully configured, generate the complete set of SWAT+ input
text files (the "TxtInOut" directory).  **This is done entirely within R**—no
external Python executable or API server is needed:

```{r write-files}
write_config_files(project)
```

By default the files are written to a `TxtInOut` sub-directory inside the
project directory.  You can customise the output location and provide a
directory of weather data files to copy:

```{r write-files-custom}
write_config_files(
  project,
  output_dir  = file.path(project$project_dir, "TxtInOut"),
  weather_dir = "/path/to/weather/data"
)
```

The function writes the full suite of SWAT+ input files including:

- **file.cio** – master index
- **time.sim** – simulation period
- **print.prt** – output configuration
- Climate station and weather generator files
- Connection, channel, reservoir, and routing files
- HRU, aquifer, basin, hydrology, and soils files
- Land-use management, operations, and structural files
- Calibration and decision-table files
- Region definition files

---

## Step 6: Run the Model

The package ships with a bundled Windows SWAT+ executable.
Use `run_swatplus()` to launch it:
  
```{r run-model}
result <- run_swatplus(
  swat_exe    = swatplus_exe(),                              
  working_dir = file.path(project$project_dir, "TxtInOut"),
  timeout     = 3600,    # 1-hour timeout
  verbose     = TRUE     # stream output to console
)
```

> **Note:** `swatplus_exe()` returns the path to the bundled Windows
> executable (`rev61.0.2_64rel.exe`).  On Linux or macOS you will need to
> supply the path to a compatible SWAT+ binary instead.

The returned list contains:

```{r run-result}
result$success  # TRUE if exit code == 0
result$elapsed  # run time in seconds
result$status   # exit code
```

---

## Complete Workflow Example

Putting it all together, here is a minimal end-to-end script:

```{r full-workflow}
library(swatplusEditoR)

# ── 1. Create the project (via rQSWATPlus) ──────────────────────────────────
dem         <- system.file("extdata", "ravn_dem.tif",      package = "rQSWATPlus")
landuse     <- system.file("extdata", "ravn_landuse.tif",  package = "rQSWATPlus")
soil        <- system.file("extdata", "ravn_soil.tif",     package = "rQSWATPlus")
lu_lookup   <- system.file("extdata", "ravn_landuse.csv",  package = "rQSWATPlus")
soil_lookup <- system.file("extdata", "ravn_soil.csv",     package = "rQSWATPlus")
outlet      <- system.file("extdata", "ravn_outlet.shp",   package = "rQSWATPlus")

project <- rQSWATPlus::qswat_run(
  project_dir       = file.path(tempdir(), "ravn_quick"),
  dem_file          = dem,
  landuse_file      = landuse,
  soil_file         = soil,
  landuse_lookup    = lu_lookup,
  soil_lookup       = soil_lookup,
  outlet_file       = outlet,
  threshold         = 500,
  slope_breaks      = c(0, 5, 15, 9999),
  landuse_threshold = 5,
  soil_threshold    = 5,
  db_file           = "swat.db",
  quiet             = TRUE
)

# ── 2. Load the project into swatplusEditoR ─────────────────────────────────
project <- load_project(project)

# ── 3. Add weather data ─────────────────────────────────────────────────────
stations <- data.frame(
  name = "ravn_met", lat = 55.85, lon = 9.65,
  pcp = "pcp.cli", tmp = "tmp.cli", slr = "slr.cli",
  hmd = "hmd.cli", wnd = "wnd.cli", pet = "pet.cli",
  atmo_dep = "atmo.cli", stringsAsFactors = FALSE
)
add_weather_stations(project, stations)
set_weather_dir(project, "/path/to/weather")

# ── 4. Configure simulation ─────────────────────────────────────────────────
set_simulation_time(project,
  day_start = 1, yrc_start = 2000,
  day_end = 365, yrc_end = 2010
)
set_print_options(project, nyskip = 2, csvout = TRUE)

# ── 5. Write input files ────────────────────────────────────────────────────
write_config_files(project)

# ── 6. Run the model ────────────────────────────────────────────────────────
result <- run_swatplus(
  swat_exe    = swatplus_exe(),
  working_dir = file.path(project$project_dir, "TxtInOut"),
  verbose     = TRUE
)

cat("Success:", result$success, "\n")
cat("Elapsed:", round(result$elapsed, 1), "seconds\n")
```

---

## Tips and Troubleshooting

* **Database access** – You can always open the project database directly for
  custom queries:

    ```r
    con <- open_project_db(project$db_file)
    DBI::dbGetQuery(con, "SELECT * FROM time_sim")
    DBI::dbDisconnect(con)
    ```

* **Parameter exploration** – To see what columns a table has, read it first:

    ```r
    con <- open_project_db(project$db_file)
    head(DBI::dbReadTable(con, "hydrology_hyd"))
    DBI::dbDisconnect(con)
    ```

* **Weather file names** – The file names in `add_weather_stations()` must
  match files present in the weather data directory or already copied to
  TxtInOut.

* **Cross-platform execution** – The bundled executable is Windows-only.  On
  Linux / macOS, compile SWAT+ from source and pass the path to
  `run_swatplus(swat_exe = "/path/to/swatplus")`.
