---
title: "Setting up a SWAT+ Simulation with swatplusEditoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up a SWAT+ Simulation with swatplusEditoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # examples require real file paths; set eval = TRUE locally
)
```

## Overview

This vignette walks through a complete SWAT+ simulation workflow using the
`swatplusEditoR` package.  It covers both ways to start a project — delineating
a new watershed from a DEM, or loading one that was already delineated in
QSWAT+ or ArcSWAT+.

The workflow mirrors the steps in the SWAT+ Editor GUI and its Python API
(`src/api/`), with one important distinction: **watershed delineation is not
part of the SWAT+ Editor Python API**.  In a typical desktop workflow,
delineation is handled by [QSWAT+](https://github.com/swat-model/QSWATPlus)
(a QGIS plugin that calls TauDEM internally); the Python API only imports the
*results* of that step.  `delineate_watershed()` brings this capability
directly into R so that the QSWAT+ dependency can be avoided entirely.

```
   Path A: Delineate watershed        Path B: Use existing watershed
   (delineate_watershed)              (use_existing_watershed)
   DEM + outlet point(s)              QSWAT+ / ArcSWAT+ shapefiles
              |                                     |
              +-------------- gis_* tables ---------+
                                    |
                         create_project_db()        SQLite project database
                         setup_project()            reference data + GIS import
                                    |
                         import_wgn()               Weather Generator data
                         import_weather()           observed weather files
                         match_weather_stations()   spatial station linkage
                                    |
                         write_swatplus_files()     all SWAT+ input text files
                                    |
                         (run SWAT+ executable)
                                    |
                         read_output()              simulation output -> SQLite
```

## Prerequisites

```r
install.packages(c("DBI", "RSQLite", "sf", "terra"))

# Install from the repository
install.packages(
  "src/r-api/swatplusEditoR",
  repos = NULL,
  type  = "source"
)
```

```{r load}
library(swatplusEditoR)
```

---

## Choosing a watershed definition method

The package provides two approaches, mirroring the SWAT+ Editor GUI:

| Option | Function | When to use |
|---|---|---|
| **Delineate watershed** | `delineate_watershed()` | You have a DEM and want the package to delineate the watershed automatically using TauDEM |
| **Use existing watershed** | `use_existing_watershed()` | You already have QSWAT+ or ArcSWAT+ output shapefiles |

Both functions produce the same result: the `gis_*` database tables are
populated and `project_config.delineation_done` is set to `TRUE`, so the rest
of the workflow is identical regardless of which path you take.

---

## Option A — Delineate watershed with TauDEM

`delineate_watershed()` uses the [TauDEM](https://hydrology.usu.edu/taudem/taudem5/) hydrological analysis library via the [`traudem`](https://github.com/lucarraro/traudem) R package.  TauDEM is the same engine used by [QSWAT+](https://github.com/swat-model/QSWATPlus) (a QGIS plugin) for watershed delineation; this function provides the equivalent capability from within R, without requiring QGIS or QSWAT+ to be installed.

### Install TauDEM and traudem

```r
# Install the R wrapper package
install.packages("traudem")

# Follow the TauDEM installation guide for your OS:
vignette("taudem-installation", package = "traudem")

# Verify everything is working
traudem::taudem_sitrep()
```

### Delineate from a DEM and outlet point

```{r delineate}
library(swatplusEditoR)
library(sf)

# Path to a projected GeoTIFF DEM
# (Geographic/lat-lon DEMs are automatically reprojected to UTM)
dem_file <- "/path/to/dem.tif"

# Outlet point – can be an sf object, data.frame(lon, lat), or shapefile path
outlet <- data.frame(lon = 174.76, lat = -36.85)   # WGS84 lon/lat

# Create the project database first
create_project_db("/path/to/project.sqlite")

# Delineate and write directly to the project database
gis_data <- delineate_watershed(
  dem               = dem_file,
  outlet            = outlet,
  project_db        = "/path/to/project.sqlite",  # optional: write to DB
  stream_threshold  = 1000,  # upstream cells to form a stream (subbasin boundary)
  channel_threshold = 200,   # optional: finer channel network within subbasins
  snap_distance     = 50,    # max cells to move outlet to nearest stream
  n_processes       = 1L,    # increase if MPI is available
  verbose           = TRUE
)
```

The function runs this TauDEM pipeline, matching the
[QSWATPlus](https://github.com/swat-model/QSWATPlus) `delineation.py`
workflow step-by-step.  The numbered progress counter reports 9 steps (11 in
dual-threshold mode); the list below includes all pipeline stages:

1. **Pit Remove** — fill sinks in the DEM
2. **D8 Flow Directions** — steepest-descent flow routing
3. **D8 Contributing Area** (full DEM)
4. **Grid Network** (`gridnet`) — path lengths (`plen`, `tlen`) and grid order (`gord`)
5. **Stream Threshold** — define stream source raster at `stream_threshold`
6. **Channel Threshold** *(optional, counted when `channel_threshold` is set)* — finer network at `channel_threshold`
7. **Move Outlets to Streams** — snap outlet to nearest channel cell
8. **D8 Contributing Area** (outlet-bounded)
9. **Grid Network** (outlet-bounded re-run, not a separate counted step)
10. **Stream Threshold** (outlet-bounded area)
11. **Channel Threshold** (outlet-bounded) *(optional)*
12. **StreamNet** — watershed raster + stream network shapefile
13. **StreamNet** (channel network) *(optional, when `channel_threshold` is set)*
14. Polygonise watershed raster → subbasin polygons (as `createWatershedShapefile()` in QSWATPlus)

The `stream_threshold` controls the **subbasin** boundaries; setting
`channel_threshold` to a smaller value adds a more detailed **channel** network
inside each subbasin — exactly how QSWATPlus uses its `numCellsSt` (stream)
and `numCellsCh` (channel) threshold parameters.

The returned `gis_data` list contains data.frames for all `gis_*` tables.
Simple HRUs (one per landscape unit, placeholder land-use `"AGRL"` and soil
`"DEFAULT"`) are created automatically.  Refine these later by importing
land-use and soil rasters via the SWAT+ Editor.

### Inspect intermediate TauDEM files

```{r delineate-inspect}
# Keep all TauDEM intermediate rasters and shapefiles for inspection
gis_data <- delineate_watershed(
  dem           = dem_file,
  outlet        = outlet,
  work_dir      = "/path/to/taudem_workspace",
  keep_work_dir = TRUE,      # don't delete intermediates
  verbose       = TRUE
)
# workspace now contains (traudem-derived names for some, explicit for others):
#   DEM.tif, DEMfel.tif                         <- pit remove
#   DEMp.tif, DEMsd8.tif                        <- d8flowdir (traudem names)
#   DEMad8.tif                                  <- aread8 (traudem names)
#   gord.tif, plen.tif, tlen.tif                <- gridnet
#   src_stream.tif, src_channel.tif             <- threshold (full DEM)
#   outlet.shp, DEMad8movedoutlet.shp           <- moveoutletstostreams
#   src_stream_final.tif, src_channel_final.tif <- threshold (outlet-bounded)
#   ord_stream.tif, tree_stream.dat, coord_stream.dat,
#   net_stream.shp, wstream.tif                 <- streamnet (streams)
#   net_channel.shp, wchannel.tif               <- streamnet (channels, if dual)
#   watershed.shp                               <- polygonised subbasins
```

---

## Option B — Use existing watershed (QSWAT+ / ArcSWAT+ output)

If you have already run QSWAT+ or ArcSWAT+ and have their output shapefiles,
use `use_existing_watershed()` to read and register them.

```{r use-existing}
# Directory produced by QSWAT+ (contains subs1.shp, rivs1.shp, etc.)
gis_dir <- "/path/to/qswatplus_output"

# Read shapefiles and write to the project database
gis_data <- use_existing_watershed(
  gis_dir     = gis_dir,
  project_db  = "/path/to/project.sqlite",   # optional: write to DB
  dem_file    = "/path/to/dem.tif",           # optional: for elevation stats
  gis_type    = "qgis",                       # "qgis" or "arcgis"
  gis_version = "3.40.3-Bratislava",          # recorded in project_config
  verbose     = TRUE
)
```

Standard QSWAT+ shapefile names recognised automatically:

| File | Contents |
|---|---|
| `subs1.shp` | Subbasins |
| `rivs1.shp` | Channels / rivers |
| `lsus2.shp` | Landscape units |
| `hrus1.shp` | HRUs |
| `reservoirs.shp` | Water bodies (optional) |
| `outlets.shp` | Point sources / inlets (optional) |
| `aquifers.shp` | Shallow aquifers (optional) |
| `deepaquifers.shp` | Deep aquifers (optional) |
| `dem.tif` | DEM raster for elevation stats (optional) |

---

## 1  Define file paths

Adjust these paths to match your local project layout.

```{r paths}
# Directory produced by QSWAT+ or ArcSWAT+ containing the delineation outputs
gis_dir      <- "/path/to/qswatplus_output"

# The SWAT+ datasets reference database (ships with the SWAT+ Editor installer)
datasets_db  <- "/path/to/swatplus_datasets.sqlite"

# Where the project database and input files will be written
project_dir  <- "/path/to/my_swat_project"

project_db   <- file.path(project_dir, "my_project.sqlite")
txtinout_dir <- file.path(project_dir, "Scenarios", "Default", "TxtInOut")
output_db    <- file.path(project_dir, "my_project_output.sqlite")

# WGN (Weather Generator) database — also ships with the SWAT+ Editor installer
wgn_db       <- "/path/to/swatplus_wgn.sqlite"

# Directory holding observed weather files (*.pcp, *.tmp, *.slr, *.hmd, *.wnd)
weather_dir  <- "/path/to/observed_weather"

# SWAT+ executable (replace with the actual version you have installed)
swat_exe     <- "/path/to/swatplus_exe"
```

---

## 2  Read GIS shapefiles and raster

`read_swatplus_gis()` scans a QSWAT+ (or ArcSWAT+) output directory for the
standard shapefile names it produces:

| Shapefile | Contents |
|---|---|
| `subs1.shp` | Subbasins |
| `rivs1.shp` | Channels / rivers |
| `lsus2.shp` | Landscape units (LSUs) |
| `hrus1.shp` | HRUs |
| `reservoirs.shp` | Reservoirs / ponds *(optional)* |
| `outlets.shp` | Point sources / inlets *(optional)* |
| `aquifers.shp` | Shallow aquifers *(optional)* |
| `deepaquifers.shp` | Deep aquifers *(optional)* |

A DEM raster is used to extract elevation statistics for each polygon when
elevation columns are absent from the shapefile attributes.

```{r read-gis}
# Read all shapefiles and the DEM raster from the QSWAT+ output directory
gis_data <- read_swatplus_gis(
  gis_dir  = gis_dir,
  dem_file = file.path(gis_dir, "dem.tif")   # optional but recommended
)

# gis_data is a named list; inspect what was found
names(gis_data)
#> [1] "subbasins"      "channels"       "lsus"           "hrus"
#> [4] "water"          "points"         "aquifers"       "deep_aquifers"

# Preview the subbasins table
head(gis_data$subbasins)
```

You can also read individual shapefiles:

```{r read-individual}
subbasins <- read_subbasins_shp(file.path(gis_dir, "subs1.shp"),
                                 dem = read_dem_raster(file.path(gis_dir, "dem.tif")))
channels  <- read_channels_shp(file.path(gis_dir, "rivs1.shp"))
lsus      <- read_lsus_shp(file.path(gis_dir, "lsus2.shp"))
hrus      <- read_hrus_shp(file.path(gis_dir, "hrus1.shp"))
```

---

## 3  Create the project database and write GIS tables

```{r create-db}
# Create the empty project database (schema only)
dir.create(project_dir, recursive = TRUE, showWarnings = FALSE)
create_project_db(project_db)

# Open a connection and write all GIS tables
con <- swat_open_db(project_db)

write_gis_to_db(
  con           = con,
  subbasins     = gis_data$subbasins,
  channels      = gis_data$channels,
  lsus          = gis_data$lsus,
  hrus          = gis_data$hrus,
  water         = gis_data$water,          # NULL if not present
  points        = gis_data$points,         # NULL if not present
  aquifers      = gis_data$aquifers,       # NULL if not present
  deep_aquifers = gis_data$deep_aquifers   # NULL if not present
)

# Quick check
swat_count(con, "gis_hrus")
#> [1] 142

swat_close_db(con)
```

---

## 4  Set up the project

`setup_project()` copies reference data (plants, soils, decision tables, land
use tables, print settings, etc.) from the SWAT+ datasets database into the
project database, and then runs the GIS import step that converts the raw GIS
tables into the full SWAT+ model object tables (routing units, channels, HRUs,
aquifers, connections, etc.).

```{r proj-setup}
setup_project(
  project_db      = project_db,
  datasets_db     = datasets_db,
  editor_version  = "2.3.0",
  project_name    = "My First R SWAT+ Project",
  is_lte          = FALSE,      # TRUE for LTE (Landscape Threshold Erosion)
  run_gis_import  = TRUE,       # convert GIS tables → SWAT+ model objects
  verbose         = TRUE
)
```

After this step, tables such as `hru_data_hru`, `hru_con`, `rout_unit_rtu`,
`chandeg_con`, `aquifer_aqu`, and the connection tables are all populated.

### Inspect what was created

```{r inspect}
con <- swat_open_db(project_db)

cat("HRUs:          ", swat_count(con, "hru_data_hru"),   "\n")
cat("Routing units: ", swat_count(con, "rout_unit_rtu"),  "\n")
cat("Channels:      ", swat_count(con, "channel_lte_cha"),"\n")
cat("Aquifers:      ", swat_count(con, "aquifer_aqu"),    "\n")

swat_close_db(con)
```

---

## 5  Import weather data

### 5a  Weather Generator (WGN)

WGN data provides synthetic weather statistics when no observed data is
available. The SWAT+ Editor ships a WGN database for the US; other regional
databases can be used by specifying `wgn_table`.

```{r wgn}
import_wgn(
  project_db = project_db,
  wgn_db     = wgn_db,
  wgn_table  = "weather_wgn_cli"
)
```

### 5b  Observed weather files

Point to a directory containing SWAT+ weather files
(`*.pcp`, `*.tmp`, `*.slr`, `*.hmd`, `*.wnd`, `*.pet`).

```{r observed-weather}
import_weather(
  project_db  = project_db,
  weather_dir = weather_dir,
  format      = "observed"   # or "2012" for SWAT 2012 format
)

# Match each weather station to the nearest spatial object (HRU, channel, etc.)
match_weather_stations(project_db)
```

### 5c  Atmospheric deposition (optional)

```{r atmo}
import_atmo_dep(
  project_db = project_db,
  csv_path   = "/path/to/atmo_dep.csv",   # columns: year, nh4_wet, no3_wet, nh4_dry, no3_dry
  filename   = "atmo.cli"
)
```

---

## 6  Configure the simulation period

The simulation time can be adjusted directly in the database or through the
SWAT+ Editor GUI. To do it programmatically:

```{r time}
con <- swat_open_db(project_db)

DBI::dbExecute(con,
  "UPDATE time_sim SET
     yrc_start = 2000,
     day_start = 0,
     yrc_end   = 2020,
     day_end   = 0,
     step      = 0")

swat_close_db(con)
```

---

## 7  Write SWAT+ input files

`write_swatplus_files()` reads every relevant table from the project database
and writes the corresponding SWAT+ plain-text input files to
`txtinout_dir`.

```{r write}
write_swatplus_files(
  project_db = project_db,
  output_dir = txtinout_dir,
  verbose    = TRUE
)
```

This creates files including `time.sim`, `print.prt`, `object.cnt`,
`codes.bsn`, `parameters.bsn`, `weather-sta.cli`, `weather-wgn.cli`,
`hru.con`, `rout_unit.con`, `chandeg.con`, `aquifer.con`, `reservoir.con`,
`hru-data.hru`, `chandeg.cha`, `aquifer.aqu`, `soils.sol`, `landuse.lum`,
`management.sch`, `lum.dtl`, and more.

---

## 8  Run the SWAT+ model

The SWAT+ executable must be run from within the `TxtInOut` directory.

```{r run-model}
# Using base R system2()
system2(swat_exe, stdout = TRUE, stderr = TRUE,
        wd = txtinout_dir)
```

Or use the `run_all()` wrapper which handles this step automatically
(see Section 9).

---

## 9  Read simulation output

After the model completes, output CSV files are written to `txtinout_dir`.
`read_output()` discovers them via `files_out.out` and imports each one as a
table in the output SQLite database.

```{r read-output}
counts <- read_output(
  project_db = project_db,
  output_db  = output_db,
  txt_dir    = txtinout_dir
)

# How many rows were imported from each output file?
print(counts)

# Query results
out_con <- swat_open_db(output_db)
swat_get_table_names(out_con)          # list all imported tables
head(swat_read_table(out_con, "hru_wb_aa"))  # annual average water balance
swat_close_db(out_con)
```

---

## 10  One-shot workflow with `run_all()`

All of the above steps can be run in a single call.  Individual stages can be
toggled on or off.

```{r run-all}
run_all(
  project_db      = project_db,
  datasets_db     = datasets_db,
  swat_exe        = swat_exe,
  output_dir      = txtinout_dir,
  output_db       = output_db,
  editor_version  = "2.3.0",
  project_name    = "My First R SWAT+ Project",

  # GIS was already imported in step 4 – skip setup + GIS re-import
  run_setup       = FALSE,
  run_gis_import  = FALSE,

  # Weather
  wgn_db          = wgn_db,
  weather_dir     = weather_dir,

  # Simulation period
  yrc_start       = 2000L,
  day_start       = 0L,
  yrc_end         = 2020L,
  day_end         = 0L,

  run_write_files = TRUE,
  run_model       = TRUE,
  run_read_output = TRUE
)
```

For a brand-new project (GIS tables already written to the database),
you can run the **complete** workflow:

```{r run-all-full}
# Full workflow from scratch
run_all(
  project_db     = project_db,
  datasets_db    = datasets_db,
  swat_exe       = swat_exe,
  output_dir     = txtinout_dir,
  output_db      = output_db,
  wgn_db         = wgn_db,
  weather_dir    = weather_dir,
  yrc_start      = 2000L,
  yrc_end        = 2020L
)
```

---

## 11  Typical project directory layout

After completing the workflow, a typical project will look like this:

```
my_swat_project/
├── my_project.sqlite          # project database (all tables)
├── my_project_output.sqlite   # simulation output tables
└── Scenarios/
    └── Default/
        └── TxtInOut/
            ├── time.sim
            ├── print.prt
            ├── object.cnt
            ├── codes.bsn
            ├── parameters.bsn
            ├── weather-sta.cli
            ├── weather-wgn.cli
            ├── hru.con
            ├── rout_unit.con
            ├── chandeg.con
            ├── hru-data.hru
            ├── chandeg.cha
            ├── aquifer.aqu
            ├── soils.sol
            ├── landuse.lum
            ├── management.sch
            ├── lum.dtl
            ├── ...
            └── (SWAT+ output: hru_wb_aa.csv, cha_wb_aa.csv, ...)
```

---

## 12  Reading individual shapefiles manually

If you need to supply a custom attribute mapping, read shapefiles individually
with the lower-level functions:

```{r manual-read}
# Read subbasins and override column name lookups
subs <- read_subbasins_shp(
  path      = file.path(gis_dir, "subs1.shp"),
  dem       = read_dem_raster(file.path(gis_dir, "dem.tif")),
  area_col  = "Area_ha",   # column name in your shapefile
  slope_col = "MeanSlope"
)

# Inspect result
str(subs)
```

---

## 13  Working with the project database directly

All project data is stored in a standard SQLite database, so you can query
and modify it with any tool that supports SQLite.

```{r direct-db}
library(DBI)

con <- swat_open_db(project_db)

# List all tables
swat_get_table_names(con)

# Read the project configuration
swat_read_table(con, "project_config")

# Check the current simulation period
swat_query(con, "SELECT * FROM time_sim")

# Update a basin parameter
swat_execute_sql(con,
  "UPDATE parameters_bsn SET esco = 0.8, epco = 0.8")

swat_close_db(con)
```

---

## See also

- `?delineate_watershed` — delineate a watershed from a DEM using TauDEM
- `?use_existing_watershed` — load QSWAT+/ArcSWAT+ shapefiles into a project
- `?run_all` — complete workflow in one function
- `?setup_project` — project initialisation and GIS import
- `?read_swatplus_gis` — read an entire QSWAT+ output directory
- `?write_swatplus_files` — write all SWAT+ input text files
- `?read_output` — import simulation output to SQLite
- [TauDEM documentation](https://hydrology.usu.edu/taudem/taudem5/)
- [traudem package](https://github.com/lucarraro/traudem)
- [SWAT+ documentation](https://swatplus.gitbook.io/docs)
- [SWAT+ Editor GitHub](https://github.com/limnotrack/swatplus-editor)
